A mixture of hidden Markov models to predict the lymphatic spread in head and neck cancer

### 2024-01-16 content/manuscript.qmd
improve mixture model formalism

Head and neck squamous cell carcinomas (HNSCC) frequently spread through the lymphatic system [@lindberg_distribution_1972;@woolgar_histological_1999]. Current diagnostic imaging modalities are unable to detect microscopic nodal metastases, which requires pathological examination of extracted tissue [@snyder_petct_2021;@strohl_petct_2021]. Since the recurrence of nodal disease is detrimental to a patient's outcome [@ho_decision_2014], large volumes in the head and neck region are often irradiated electively to minimize the risk of missing occult disease. Decision guidelines about which nodal regions -- i.e., anatomically defined lymph node levels (LNLs) -- to irradiate [@vorwerk_guidelines_2011;@biau_selection_2019] are currently not based on a patient's individual risk, but only on the overall prevalence of nodal disease as reported in the literature [@lindberg_distribution_1972;@woolgar_histological_1999].

@@

@@

![Parametrized graph representation of the lymphatic network considering four LNLs. Blue, round nodes represent the hidden RVs, while orange square nodes show the observed RVs. Arcs represent a conditional probability parametrized with the quantity noted next to it.](../static/graph_with_obs.svg){#fig-graph-with-obs width="25%"}

@@

To train our model, we need to compute the likelihood of a dataset $\mathcal{D}=\left( \mathbf{y}_1, \mathbf{y}_2, \ldots, \mathbf{y}_N \right)$ of $N$ previously observed patient diagnoses, given the model's parameters $\boldsymbol{\theta}$. This is given by a product of the sums in the denominator of @eq-bayes-law:

$$
\mathcal{L}(\mathcal{D} \mid \boldsymbol{\theta}) = \prod_{i=1}^N \underbrace{\sum_{j=1}^{2^V} P \left( \mathbf{Y} = \mathbf{y}_i \mid \mathbf{X} = \boldsymbol{\xi}_j \right) P \left( \mathbf{X} = \boldsymbol{\xi}_j \mid T_i, \boldsymbol{\theta} \right)}_{P \left( \mathbf{Y} = \mathbf{y}_i \mid T_i, \boldsymbol{\theta} \right)}
$$

@@

In such cases, mixture models are often considered. They assume the data to come from a finite mixture distribution, which -- in our particular case -- can be written following @mclachlan_finite_2019:

$$
P \left( \mathbf{Y} \mid \boldsymbol{\Psi} \right) = \sum_{i=0}^g \pi_i \cdot P \left( \mathbf{Y} \mid \boldsymbol{\theta}_i \right)
$$ {#eq-mixture-distribution}

Here, the $\pi_i \in [0,1]$ are the mixing proportions with $\sum_{i=0}^g \pi_i = 1$, while the $\boldsymbol{\Psi} = \left( \boldsymbol{\theta}_1, \ldots, \boldsymbol{\theta}_g, \pi_1, \ldots, \pi_g \right)$ is the entire vector of unknown parameters in this mixture of $g$ components. Note that we will implement our model such that some of the parameters in each $\boldsymbol{\theta}_i$ are shared across the $g$ components -- namely the $t_{r \rightarrow v}$ corresponding to the blue arcs in @fig-graph-with-obs.

Let now $\mathcal{D} = \left( \mathcal{D}_1, \ldots, \mathcal{D}_s \right)$ be a dataset consisting of $s$ subcohorts each corresponding to patients with tumor in the same subsite. Also, let subcohort $i$ have $N_i$ patients in it.

We can now think of a way of generating this dataset $\mathcal{D}$ under this mixture model:

1. We draw a categorical RV $C_j$ that may take on the values $1, \ldots, g$ with corresponding probabilities $\pi_1, \ldots, \pi_g$. Hence, we have $P(C_j=c_j) = \pi_{c_j}$. For convenience, we also transform this into a one-hot encoding $\mathbf{z}_j$ that is zero everywhere, except $z_{\ell j}=1$ when $C_j=\ell$:
  $$
  P \left( \mathbf{Z}_j=\mathbf{z}_j \right) = \prod_{i=0}^g \pi_i^{z_{ij}}
  $$

2. Then, we draw $\mathbf{y}_j$ from the conditional probability of $P \left( \mathbf{Y}_j \mid \mathbf{Z}_j, \boldsymbol{\Psi} \right)$, which is now given by a slightly modified version of @eq-mixture-distribution:
  $$
  P \left( \mathbf{Y}_j \mid \mathbf{Z}_j, \boldsymbol{\Psi} \right) = \frac{\sum_{i=0}^g z_{ij} \pi_i P \left( \mathbf{Y}_j \mid \boldsymbol{\theta}_i \right)}{\sum_{i=0}^g z_{ij} \pi_i} = P \left( \mathbf{Y}_j \mid \boldsymbol{\theta}_\ell \right)
  $$
  where $\ell$ is just so that $z_{\ell j}=1$, and all other entries of $\mathbf{z}_j$ are zero.

@@

# Discussion

>>>>

Head and neck squamous cell carcinomas (HNSCC) frequently spread through the lymphatic system [@lindberg_distribution_1972;@woolgar_histological_1999]. Current diagnostic imaging modalities are unable to detect microscopic nodal metastases, which requires pathological examination of extracted tissue [@snyder_petct_2021;@strohl_petct_2021]. Since the recurrence of nodal disease is detrimental to a patient's outcome [@ho_decision_2014], large volumes in the head and neck region are often irradiated electively to minimize the risk of missing occult disease. Decision guidelines about which nodal regions -- i.e., anatomically defined lymph node levels (LNLs) -- to irradiate [@biau_selection_2019] are currently not based on a patient's individual risk, but only on the overall prevalence of nodal disease as reported in the literature [@lindberg_distribution_1972;@woolgar_histological_1999].

@@

![Parametrized graph representation of the lymphatic network considering four LNLs. Blue, round nodes represent the hidden RVs, while orange square nodes show the observed RVs. Arcs represent a conditional probability parametrized with the quantity noted next to it.](../static/graph_with_obs.svg){#fig-graph-with-obs width="22%"}

@@

@@

To train our model, we need to compute the likelihood of a dataset $\mathbf{D}=\left( \mathbf{y}_1, \mathbf{y}_2, \ldots, \mathbf{y}_N \right)$ of $N$ patients, given the model's parameters $\boldsymbol{\theta}$. This is given by a product of the sums in the denominator of @eq-bayes-law:

$$
P \left(\mathbf{D} \mid \boldsymbol{\theta} \right) = \prod_{i=1}^N \underbrace{\sum_{j=1}^{2^V} P \left( \mathbf{Y} = \mathbf{y}_i \mid \mathbf{X} = \boldsymbol{\xi}_j \right) P \left( \mathbf{X} = \boldsymbol{\xi}_j \mid T_i, \boldsymbol{\theta} \right)}_{P \left( \mathbf{Y} = \mathbf{y}_i \mid T_i, \boldsymbol{\theta} \right)}
$$ {#eq-likelihood}

@@

In such cases, mixture models are often considered. They assume the data to come from a finite mixture distribution, which -- in our particular case -- can be written as follows:

$$
P \left( \mathbf{D} \mid \boldsymbol{\Psi} \right) = \sum_{j=0}^g c_j P \left( \mathbf{D} \mid \boldsymbol{\theta}_j \right)
$$ {#eq-mixture-distribution}

Here, the $\mathbf{c} \in [0,1]^g$ is the vector of mixing proportions with $\sum_{j=0}^g c_j = 1$, while the $\boldsymbol{\Psi} = \left( \boldsymbol{\theta}_1, \ldots, \boldsymbol{\theta}_g\right)$ is the vector of all $g$ models' parameters. Note that we will implement our model such that some of the parameters in each $\boldsymbol{\theta}_j$ are shared across the $g$ components -- namely the $t_{r \rightarrow v}$ corresponding to the blue arcs in @fig-graph-with-obs.

Let now $\mathfrak{D} = \left( \mathbf{D}_1, \ldots, \mathbf{D}_s \right)$ be a dataset consisting of $s$ subcohorts of patients. Within a subcohort $i$ we find $N_i$ patients with tumors in the same subsite. We can then introduce a _latent variable_ $\mathbf{Z}$ with a one-hot-encoding: Basically, it can take on values $\mathbf{z}_i \in \{ 0,1 \}^g$ with $z_{ij}=1$ if subcohort $i$ belongs to component $j$ and $z_{ij}=0$ else.

Introducing this latent variable is helpful, because the mixture model likelihood is invariant w.r.t. permutations of the component labels. This introduces a number of problems, e.g. for common MCMC sampling methods. The latent variable $\mathbf{Z}$ allows us to derive two sets of interdependent equations that we may solve in an iterative fashion (see e.g. @bishop_pattern_2006 for a detailed derivation) that is commonly referred to as _expectation-maximization (EM)_ algorithm:

The first are the probabilities of subcohort $i$ to belong to component $j$, given a set of parameters $\boldsymbol{\Psi}^\star$. These are often called the _responsibilities_:
$$
\gamma (z_{ij}) = P \left( z_{ij}=1 \mid \mathbf{D}_i , \boldsymbol{\Psi}^\star, \mathbf{c} \right) = \frac{ c_j P \left( \mathbf{D}_i \mid \boldsymbol{\theta}_j^\star \right) }{\sum_{k=0}^g c_k P \left( \mathbf{D}_i \mid \boldsymbol{\theta}_k^\star \right)}
$$

From this, we can compute new mixing proportions $c_j^\star=\sum_{i=1}^s \gamma (z_{ij}) / s$ and then infer new parameters $\boldsymbol{\Psi}^\star$ -- e.g. via MCMC sampling -- from the resulting likelihood:

$$
P \left( \mathbf{D} \mid \boldsymbol{\Psi}, \mathbf{c}^\star \right) = \sum_{j=0}^g c_j^\star P \left( \mathbf{D} \mid \boldsymbol{\theta}_j \right)
$$

## Implementation {#sec-implementation}

We did this in this and that fashion...

@@

Works super well, of course!

# Discussion {#sec-discussion}

All that's left is for other people to bring this into clinical practice.
